---
title: "How to choose how to model"
output: 
  bookdown::html_document2:
    base_format: rmarkdown::html_vignette
    fig_caption: yes
bibliography: "covid.bib"
link-citations: yes
vignette: >
  %\VignetteIndexEntry{Priors exploration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

4d-Modeller and INLA/inlabru allow you to use both fixed and random effects in your model with SPDEs allowing you to model continuous random processes in your data. Below we explain briefly each of these processes and how to choose when to use them.

In this document we will use the following data for examples:

```{r, error=TRUE}
library(MASS)  # for true multivariate normal simulation
set.seed(123)

n <- 100  # number of observations
timesteps <- 10  # number of time steps

# Create a basic time and space grid
df <- expand.grid(x = seq(1, 10, length.out = sqrt(n)),
                  y = seq(1, 10, length.out = sqrt(n)),
                  time = 1:timesteps)

# Create a covariance matrix
cov_matrix <- matrix(0.5, nrow = n, ncol = n)  # Example covariance matrix with 0.5 correlation
diag(cov_matrix) <- 1  # Diagonal should be 1

# Generate spatially and temporally correlated noise for each time step
for (i in unique(df$time)) {
  noise <- mvrnorm(n = 1, mu = rep(0, n), Sigma = cov_matrix)
  df$value[df$time == i] <- noise
}

head(df)

library(ggplot2)

ggplot(df, aes(x = x, y = y, color = value)) +
  geom_point() +
  facet_wrap(~ time) +
  scale_color_viridis_c() +
  theme_minimal()


```

## Fixed Effects

A fixed effect in a bayesian model is an effect whose value is fixed in a location but has a probability distribution associated with it's uncertainty. A fixed effect represents a parameter or coefficient in a model that is assumed to have a fixed, constant value across different observations/locations. In a Bayesian model, a fixed effect is a special case of a random effect where we assume the effect has zero variance.

### why would you choose a fixed effect?

You would choose to model an effect as a fixed effect if you believe the effect will always have the same outcome no matter the context of that the observation is taken. For example, a unit of rain should fill a reservoir by that an equivalent unit no matter the conditions given constant climatic conditions.

### example of a fixed effect in r-inla/inlabru

```{r, error=TRUE}
example code
```

## Random Effects

A random effect in a bayesian model has an effect whose value is drawn from a probability distribution and has associated uncertainty drawn from it's own distribution.  


```{r, error=TRUE}
# Define a formula for the model
# Load the INLA library
library(INLA)

# Define the formula for the INLA model
formula <- value ~ 1 + f(x, y, model = "iid") + f(time, model = "rw1")

# Fit the INLA model
result <- inla(formula, data = df, family = "gaussian")

# Print the summary of the INLA model
summary(result)
```

### Random Effects of Unobserved Processes

```{r, error=TRUE}
example code
```

### Random Effects Constrained by Observed Processes

```{r, error=TRUE}
# Define the model
formula <- value ~ f(space, model = "rw2") + f(time, model = "rw1")

# Fit the model using INLA
result <- inla(formula, data = df, family = "gaussian")

# Summary of the model
summary(result)
```

### Comparing the output

Finally, we can compare the output of all the models using the DIC. This is important as we would like to know the fittedness of each model even if we have some theoretical belief that a model should be specificied in a certain way.