---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Installation

To get the 4DModeller R package `fdmr` installed first you need to make sure you have a recent version of R installed.
The easiest way to do this is to [install RStudio](https://posit.co/downloads/).

As `fdmr` has not yet been submitted to [CRAN](https://cran.r-project.org/) we'll need to install it from our GitHub repository.
We'll use the `devtools` package to do this so if you don't have it installed already in an R session run

```{r eval=FALSE}
install.packages("devtools")
library(devtools)
```

Now we're ready to install `fdmr`

```{r eval=FALSE}
devtools::install_github("4DModeller/fdmr")
```

You should now have `fdmr` and all its dependencies installed and you can continue on [one of our tutorials](https://4dmodeller.github.io/fdmr/articles/).

## Installation issues

Here we cover issues we've encountered getting `fdmr` installed on different platforms.

### Linux host and C++17 error

Users on Linux systems may encounter issues during compilation of packages using C++17. You may see an error similar to

```
'Error: C++17 standard requested but CXX17 is not defined'
```

To fix this we need to tell R to use a compiler that supports C++17. Here we tell it to use `g++` and pass in the `-std=gnu++17` flag to enable C++17 support.

```{bash eval=FALSE}
mkdir -p ~/.R
echo 'CXX17 = g++ -std=gnu++17 -fPIC' > ~/.R/Makevars
```

After running these commands try reinstalling `fdmr` and the packages should now compile correctly.
