---
title: "Plotting maps"
output: html_notebook
---

4DModeller provides some simple map plotting functions that help you create maps from many different types of data.

## Polygons

We'll first start off by plotting some polygon data on a map. To do this we'll use some of the data we use in our Hydrology tutorial, where we display a catchment area on map.
First we need to retrieve the example data, then load in the raster data from file.

```{r}
fdmr::retrieve_tutorial_data(dataset = "hydro")
norway_polygon_location <- fdmr::get_tutorial_datapath(filename = "Kvilldal_Catch_Boundary.geojson")
```

Next we need to manipulate the data to ensure we plot the area correctly.

```{r}
norway_polygon <- rgdal::readOGR(norway_polygon_location)
norway_polygon <- sf::st_as_sf(norway_polygon
                               , coords=c("longitude", "latitude")
                               , crs="+proj=utm +zone=32")

sfc <- sf::st_transform(norway_polygon, crs = "+proj=longlat +datum=WGS84")

leaflet::leaflet(sfc, height = 500, width = 750) %>%
  leaflet::addTiles("https://server.arcgisonline.com/ArcGIS/rest/services/World_Imagery/MapServer/tile/{z}/{y}/{x}",
           attribution = 'Tiles &copy; Esri &mdash; Source: Esri, i-cubed, USDA, USGS, AEX, GeoEye, Getmapping, Aerogrid, IGN, IGP, UPR-EGP, and the GIS User Community') %>%
  leaflet::addPolygons(stroke = FALSE, smoothFactor = 0.3,
              fillColor = "red",
              fillOpacity = 0.5)
```

> **_NOTE:_**  These functions are still a work in progress and will probably be combined into a single plot_map function once we know what interface we want


```{r eval=FALSE}
fdmr::plot_map_polygon(data=sfc, tiles="satellite")
```

## Rasters

> **_NOTE:_**  This will be converted to use the `fdmr::plot_map` function, or maybe `fdmr::plot_interactive_map` as we want to select from the stack.

Here we load in a stack of rasters from a NetCDF file. We do this using the `raster::stack` function which creates a `RasterStack` object. We'll also have a quick look at the data object.

```{r}
era5_data_filepath <- fdmr::get_tutorial_datapath(filename = "era5_land_daily.nc")
era5_precip <- raster::stack(era5_data_filepath)
era5_precip
```
We can see that we've got a stack of 31 layers, one for each day of October 2021. To list all the dates in the stack we can do

```{r}
names(era5_precip)
```

As we want to plot this on a Leaflet map we need to ensure the coordinate reference system (CRS) is correct.

```{r}
projected_raster <- raster::projectRaster(era5_precip, crs = "+proj=utm +zone=32")
```

As we only want to plot the data that overlaps with our rain catchment area polygon we plotted earlier, we mask the data using `raster::mask` and then crop it using `terra::crop`.

```{r}
crop_era5 <- raster::mask(projected_raster, norway_polygon)
crop_era5 <- terra::crop(crop_era5, raster::extent(norway_polygon), snap="near")
```

We're now read to plot the map. For the moment we only have the functionality to plot a single raster from the stack, but we're working on a `plot_interactive_map` function that will allow the dynamic selection of layers from the stack using a date slider.

For now we'll select the layer containing data for the 1st of October 2021 
```{r}
raster_image <- crop_era5$X2021.10.01
```

Now we're ready to plot the data, using the same `sfc` object we created above.

```{r}
leaflet::leaflet(sfc, width=750, height=500) %>% 
  leaflet::addTiles() %>%
  leaflet::addRasterImage(x = raster_image, 
                 opacity = 1) %>%
  leaflet::addPolygons(stroke = FALSE, smoothFactor = 0.3,
              fillColor = "red",
              fillOpacity = 0.5)
```

> **_NOTE:_**  These functions are still a work in progress and will probably be combined into a single plot_map function once we know what interface we want

We can now plot the raster data using the `plot_map_raster` function.

```{r}
fdmr::plot_map_raster(raster_data=raster_image, polygon_data=sfc)
```


## Model output

We can also plot the output of models on a map ...

- What object do we expect back from the model?


