---
title: "Map Plotting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Map Plotting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

4DModeller provides some simple map plotting functions that help you create maps from many different types of data.

```{r}
library(fdmr)
```
## Polygons

We'll first start off by plotting some polygon data on a map. To do this we'll use some of the data we use in our Hydrology tutorial, where we display a catchment area on map.
First we need to retrieve the example data, then load in the raster data from file.

```{r}
fdmr::retrieve_tutorial_data(dataset = "hydro")
norway_polygon_location <- fdmr::get_tutorial_datapath(filename = "Kvilldal_Catch_Boundary.geojson")
```

Next we need to manipulate the data to ensure we plot the area correctly,

```{r}
norway_polygon <- rgdal::readOGR(norway_polygon_location)
norway_polygon <- sf::st_as_sf(norway_polygon,
  coords = c("longitude", "latitude"),
  crs = "+proj=utm +zone=32"
)

sfc <- sf::st_transform(norway_polygon, crs = "+proj=longlat +datum=WGS84")
```

Now we can plot the polygon on a leaflet map using `plot_map`.

```{r}
fdmr::plot_map(polygon_data = sfc)
```

## Rasters

We can also plot raster data onto the map. We'll now read some raster data from a [NetCDF](https://www.unidata.ucar.edu/software/netcdf/) file. We do this using the `raster::stack` function which creates a `RasterStack` object which itself contains a number of `RasterLayer` objects. These hold [ERA5](https://www.ecmwf.int/en/forecasts/dataset/ecmwf-reanalysis-v5) precipitation data for each day of the month for October 2021.

```{r}
era5_data_filepath <- fdmr::get_tutorial_datapath(filename = "era5_land_daily.nc")
era5_precip <- raster::stack(era5_data_filepath)
era5_precip
```
We can see that we've got a stack of 31 layers, one for each day of October. To list all the dates in the stack we can do

```{r}
names(era5_precip)
```

As we want to plot this on a Leaflet map we need to ensure the coordinate reference system (CRS) is correct.

```{r}
projected_raster <- raster::projectRaster(era5_precip, crs = "+proj=utm +zone=32")
```

As we only want to plot the data that overlaps with our rain catchment area polygon we plotted earlier, we mask the data using [`raster::mask`](https://www.rdocumentation.org/packages/raster/versions/3.6-20/topics/mask) and then crop it using [`terra::crop`](https://www.rdocumentation.org/packages/terra/versions/1.7-29/topics/crop).

```{r}
crop_era5 <- raster::mask(projected_raster, norway_polygon)
crop_era5 <- terra::crop(crop_era5, raster::extent(norway_polygon), snap = "near")
```

We're now read to plot the raster data. Let's select a single layer from the stack, the first day of October, and plot the precipitation measurements for that day over the polygon we previously plotted.

```{r}
raster_image <- crop_era5$X2021.10.01
```

Now we're ready to plot both the polygon and raster data.

```{r}
fdmr::plot_map(raster_data = raster_image, polygon_data = sfc, palette = "viridis")
```

## Interactive mapping

You can plot data on an interactive [`leaflet`](https://rstudio.github.io/leaflet/) map using our `plot_interactive_map` function. This launches a [Shiny](https://shiny.posit.co/) app allowing you to customise your plot to better
investigate your data. Below are a couple of examples showing how to use the interactive plotting tool.

### Hydroelectric Power

We want to create a plot on the map covering the rain catchment area around the power plant. To do this we read in a GeoJSON file containing the coordinates of the catchment area and make sure it's in the correct coordinate reference system to plot on our map.

```{r}
norway_polygon_location <- fdmr::get_tutorial_datapath(filename = "Kvilldal_Catch_Boundary.geojson")
norway_polygon <- rgdal::readOGR(norway_polygon_location)
norway_polygon <- sf::st_as_sf(norway_polygon,
  coords = c("longitude", "latitude"),
  crs = "+proj=utm +zone=32"
)

norway_polygon <- sf::st_transform(norway_polygon, crs = "+proj=longlat +datum=WGS84")
```

Next we read ECMWF Reanalysis v5 (ERA5) precipitation data from a NetCDF file.

```{r}
era5_data_filepath <- fdmr::get_tutorial_datapath(filename = "era5_land_daily.nc")
era5_precip <- raster::stack(era5_data_filepath)
era5_precip <- raster::projectRaster(era5_precip, crs = "+proj=utm +zone=32")
```

The data is now ready to pass to our interactive plotting function.

```{r eval=FALSE}
fdmr::plot_interactive_map(raster_data = era5_precip, polygon_data = norway_polygon)
```
