---
title: "Mesh building"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hydrology}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Here we'll cover building a mesh from some spatial data, modifying its parameters and plotting it and the data on an interactive map.

We'll first load in some spatial data to create the mesh. This is the same data we previously used in the COVID-19 tutorial.

```{r}
sp_data <- fdmr::load_tutorial_data(filename = "spatial_data.rds")
```

Let's start the mesh builder app and pass it a selection of the spatial data we just loaded. We'll select just the longitude and latitude data.
This data will also get 

```{r}
# loc_data = sp_data@data[, c("LONG", "LAT")]
```

```{r}
fdmr::mesh_builder(spatial_data=sp_data)
```

> **_NOTE:** This section will be moved into the mesh_builder app

We first get the CRS of the spatial data.

```{r}
crs <- sf::st_crs(sp_data)
crs
```

Now we're ready to build the mesh

```{r}
initial_range <- diff(range(sp_data@data[, "LONG"])) / 5

max_edge <- initial_range / 8

mesh <- INLA::inla.mesh.2d(
  loc = sp_data@data[, c("LONG", "LAT")],
  max.edge = c(1, 2) * max_edge,
  offset = c(initial_range / 4, initial_range),
  cutoff = max_edge / 7,
  crs = crs,
)
```

Now we've built the mesh let's convert it to an object we can plot using `leaflet`. This gives us a list of vertices and points.

```{r}
polygon_data <- fdmr::mesh_to_spatial(mesh)
```

Let's just take the triangles first, this is a `SpatialPointDataFrame` that contains the mesh triangles.

```{r}
triangles <- polygon_data$triangles
triangles
```

Now we can create a simple `leaflet` plot. As the mesh is quite large it may take a short while to create the plot.

```{r}
leaflet::leaflet(triangles) %>% leaflet::addTiles() %>% leaflet::addPolygons(weight=1, fillOpacity = 0.2)
```