---
title: "Simulation Tutorial (Gaussian data)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation Tutorial (Gaussian data)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In this tutorial, we'll start by generating some simulated spatio-temporal data, with the city of Bristol as our study region. Next, we'll walk you through the steps of running the Bayesian Hierarchical Model (BHM) using the INLA-SPDE approach. Finally, we'll demonstrate the use the key functionality provided by the `fdmr` package.

# Data simulation

First we simulate some Gaussian distributed data observed at 55 spatial locations in Bristol, UK and across 6 time points. We use the `fdmr::retrieve_tutorial_data` function and the `fdmr::load_tutorial_data` function to retrieve and load in the geographical information for the spatial locations in Bristol from the `fdmr` example data store.

```{r loadsp}
fdmr::retrieve_tutorial_data(dataset = "priors")
sp_data <- fdmr::load_tutorial_data(dataset = "priors", filename = "spatial_dataBris.rds")
```

Then we simulate the Gaussian data that exhibit spatio-temporal correlation. These simulated data are finally organized into a data frame named `simdf`, where each row corresponds to an observation at a specific location and time. The variable `ID` is the spatial location identifier. Variable `time` denotes the time point. The variable `datn` represents the simulated response data. The longitude and latitude coordinates for each location are stored in the variables `LONG` and `LAT`, respectively.

```{r simdat} 
library(INLA) 

n.time<-6
n <- 55
locs <- as.matrix(sp_data@data[,c("LONG","LAT")])

initial_range <-diff(range(locs[,1]))/3
max_edge<-initial_range /2
mesh <- fmesher::fm_mesh_2d(loc=locs,
                           max.edge = c(1,2)*max_edge,
                           offset=c(initial_range , initial_range ),
                           cutoff = max_edge/7
)


spde <- INLA::inla.spde2.pcmatern(
  mesh = mesh, 
  prior.range = c(initial_range , 0.5), 
  prior.sigma = c(1, 0.01) 
)

sigma0 <- 7
range0 <- 0.03
kappa0 <- sqrt(8 / 1) / range0
tau0 <- 1 / (sqrt(4 * pi) * kappa0 * sigma0) 
inla.seed <- sample.int(n=1E4, size=n.time)

Q <- INLA::inla.spde.precision(spde, theta = c(log(tau0), log(kappa0)))
x.mat<-matrix(NA,ncol=n.time,nrow =mesh$n)
for (co in 1:ncol(x.mat)){
  x.mat[,co]<-INLA::inla.qsample(n = 1, Q)
} 

A <- INLA::inla.spde.make.A(mesh=mesh, loc=locs)
x.dat<-matrix(NA,ncol=n.time,nrow =n)
for (t in 1:ncol(x.dat)){
  x.dat[,t] <- drop(A %*% x.mat[,t]) 
}


alpha<-0.9
sp.mat<-matrix(NA,ncol = n.time,nrow = n)
sp.mat[,1]<-x.dat[,1]
for (t in 2:n.time){
  sp.mat[,t]<-alpha*sp.mat[,t-1]+x.dat[,t]
}


beta0 <- 0.5 
sigma_e <- 0.1 
lin.pred <- beta0 + sp.mat
y.mat <- lin.pred+matrix(rnorm(n*n.time,0,sigma_e),ncol=n.time)

simdf<-data.frame(ID=rep(1:n,each=n.time),
                  time = rep(c(1:n.time),n),
                  datn = as.numeric(t(y.mat)), 
                  LONG = rep(locs[,1],each=n.time),
                  LAT = rep(locs[,2],each=n.time))

```

A histogram is created to visualize the distribution of the simulated Gaussian data.

```{r hist,error=TRUE,fig.cap="A histogram of the simulated Gaussian data.", fig.width=8, fig.height=4, fig.align = "center"}
hist(simdf$datn,main='', xlab='Data')
```

We can also map the simulated data using the `fdmr::plot_map` function. For example, a map of the simulated data at the first time point is given as

```{r mapsimdat,error=TRUE,fig.cap="A map of the simulated data in Bristol at the first time point.", fig.width=8, fig.height=4, fig.align = "center"}
fdmr::plot_map(
  polygon_data = sp_data,
  domain = simdf[simdf$time==1,"datn"],
  palette = "Reds",
  legend_title = "Data",
  add_scale_bar = TRUE,
  polygon_fill_opacity = 1,
  polygon_line_colour = "transparent"
)

```


# Model fitting

With the simulation dataset now prepared, our next step is to implement a Bayesian spatio-temporal model using the INLA-SPDE approach.

## Generate mesh, build the SPDE model, and set priors for the spatial parameters

```{r meshSPDE}
locs <- unique(simdf[, c("LONG", "LAT")])
initial_range <-diff(range(locs[,1]))/3
max_edge<-initial_range /2
mesh <- fmesher::fm_mesh_2d(loc=locs,
                           max.edge = c(1,2)*max_edge,
                           offset=c(initial_range , initial_range ),
                           cutoff = max_edge/5
)

spde <- INLA::inla.spde2.pcmatern(
  mesh = mesh, 
  prior.range = c(initial_range , 0.5), 
  prior.sigma = c(1, 0.01) 
)

```

## Define how the process evolves over time, set prior for the temporal parameter and define a temporal index

```{r pccor, error=TRUE}
rhoprior <- base::list(theta = list(prior = 'pccor1', param = c(0, 0.9)))
group.index<-simdf$time
n_groups<-length(unique(simdf$time)) 
```

We will use the function `bru()` of package `inlabru` to fit the model. `bru` expects the coordinates of the data, thus we transform `simdf` data set to a SpatialPointsDataFrame using the function `coordinates()` of the `sp` package.

```{r, error=TRUE}
sp::coordinates(simdf)<- c("LONG","LAT")
```

## Define the model formula
```{r formula, error=TRUE}
 formula <- datn ~ 0 + Intercept(1) +
   f(main=coordinates,  
     model=spde,
     group=group.index,
     ngroup=n_groups,
     control.group=list(model="ar1",
                        hyper=rhoprior))

```


## Fit the model

Finally, we fit the spatio-temporal model using SPDE approach and AR(1) process by calling the function `bru()` of the package `inlabru`.

```{r fitmodel}
inlabru_model <- inlabru::bru(formula,
  data = simdf,
  family = "gaussian",
  options = list(
    verbose = TRUE
  )
)

```

## Results summary

After completing the model fitting process, you can examine the main result summaries by typing `summary(inlabru_model)`. Additionally, you can check the parameter estimates for fixed effects, random effects, and hyperparameters as follows.

```{r summary,eval=FALSE}
model_summary<-summary(inlabru_model)
model_fixed <- inlabru_model$summary.fixed
model_random<-inlabru_model$summary.random
model_hyperparams <- inlabru_model$summary.hyperpar
```


The Deviance Information Criterion (DIC) value, which measures the goodness of model fit, can be examined as follows.


```{r dicvalue}
model_dic<-inlabru_model$dic$dic

```


You can obtain the model's fitted values and compare them with the observed values as follows.

```{r fittedvals,eval=TRUE}

fitted_vals<-inlabru_model$summary.fitted.values[1:nrow(simdf),]

```

```{r plotfitted, error=TRUE,fig.cap="A plot of the fitted and observed values.", fig.width=8, fig.height=4, fig.align = "center"}
par(pty="s")
plot(fitted_vals$mean,simdf$datn, xlab='fitted',ylab='observed')

```

# Exploring the functionality of the `fdmr` Package

Next, we'll demonstrate the key functionality offered by the `fdmr` package.

## fdmr::plot_mesh 

The `fdmr::plot_mesh` function displays both the mesh and the observation points overlaid on the mesh.


```{r meshplot, error=TRUE,fig.cap="A plot of the mesh.", fig.width=8, fig.height=4, fig.align = "center"}

fdmr::plot_mesh(mesh=mesh, spatial_data=simdf@coords)
```

## fdmr::mesh_builder

`fdmr` provides an interactive mesh builder tool called `fdmr::mesh_builder`. It allows you to build a mesh from some spatial data, modify its parameters and plot it and the data on an interactive map. To build a mesh, we need to pass `sp_data` to the `mesh_builder` function.

```{r meshbuilder,eval=FALSE}
fdmr::mesh_builder(spatial_data = sp_data)
```

The interactive map allows you to customise the initial parameters such as `max_edge`, `offset` and `cutoff` to change the shape and size of the mesh. We also provide a simple function to check if there are any errors with the meshes created using the mesh builder tool. To use the mesh checking functionality you must pass your measuremnet data to the `fdmr::mesh_builder` function. Specifically, create a mesh of your design using the meshbuilder tool. When you’re finished, click the “Check mesh” button on the user interface. This passes the created mesh to the `fdmr::mesh_checker` function and returns a list containing any errors found with the mesh.

```{r meshchecking,eval=FALSE}
fdmr::mesh_builder(spatial_data = sp_data, obs_data = simdf)
```

## Interactive priors Shiny app
`fdmr` provides a Priors Shiny app which allows you to interactively set and see the model fitting results of different priors. You can launch this app by passing the spatial and observation data to the `fdmr::model_builder` function.

```{r priorapp, eval=FALSE}
fdmr::model_builder(spatial_data = sp_data, 
                         measurement_data = simdf, 
                         mesh=mesh, 
                         time_variable="time")
```

## fdmr::plot_map

The `fdmr::plot_map` function generates a map of the region based on the provided SpatialPolygonDataFrame.

```{r rmap, error=TRUE,fig.cap="A map of the study region.", fig.width=8, fig.height=4, fig.align = "center"}
fdmr::plot_map(polygon_data = sp_data)
```
