---
editor_options: 
  markdown: 
    wrap: 72
---

```{r modules}
library(ncdf4)
library(tidyverse)
```

# Import and load the data

```{r dataload}
# specify where your data is stored locally
ncpath <- "N:/courses_conferences/2023_4DM_hackathon/fdmr/vignettes/data/" 
ncname <- "Bayelva_snow" 
ncfname <- paste(ncpath, ncname, ".nc", sep="")

#open data 
ncin <- nc_open(ncfname) 
print(ncin)
```

# Get relevant features

The netcdf file has both point measurements of snow depth (ds, or
translated into SWE: Ds) for several years, and continuous raster data
with snow depth duration for several years and topographic features. We
start with the snow depth points only:

```{r dataselect}
# specify the longitude and latitude columns. Note that the coordinates are utm!
longitude <- ncvar_get(ncin,"xs") 
latitude <-ncvar_get(ncin,"ys")

# Get the first year of snow depth (ds)
ds <- ncvar_get(ncin,"ds")[,1]

# Create a data frame
df <- data.frame(LONG = longitude,LAT = latitude,ds) 
head(df)
```

# Build the mesh

The shiny app doesn't work with UTM, so we need to do this
non-interactively.

```{r plotmesh}
# This doesn't work with UTM:
# fdmr::mesh_builder(spatial_data = df, crs="+proj=utm +zone=33")

# We have to define an UTM
crs <- "+proj=utm +zone=33" 
mesh <- fmesher::fm_mesh_2d_inla(loc =
df[,c(1,2)], max.edge = c(20,40),crs = crs ) 

plot(mesh)
```

# Visualise the mesh: also plot our data points

This doesn't seem to work with the built-in plot_mesh command in UTM.

```{r plotmeshandpoints}

fdmr::plot_mesh(mesh=mesh, spatial_data = df[,c(1,2)])
```
